DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-08-26 2015CompleteDerivation/completeDerivation.tex
Another pass for quantum prelude

As we saw in the previous sections, classical particles should not be understood as point-like finite mass objects as this would open a series of problems and inconsistencies. The issues created by such misidentification are only amplified within the context of quantum mechanics. In fact, many issues preventing a satisfying intuitive picture stem from this confusion.

The problem goes along these line. One assumes, implicitly or explicitly, that quantum particles are point-like. But they don't always behave like one, they also behave like a wave, a distribution. During measurements, it coincides experimentally with a probability distribution. So the state of a quantum particles is a probability distribution. Except that it self-interacts. So its a weird quantum probability distribution. The quantum particle behaves like its in more than one place at once. We learn this picture and, sometimes begrudgingly, we get used to it.

Here we contend that this picture is not only confusing and unnecessary, it also prevents to understand fully the deep links between classical and quantum mechanics, in the same way that thinking of classical particles as point-like objects prevented to see the holistic picture in classical mechanics.

The first thing to note here is that all objects behave like they are in more than one place at once. The earth is both under Milano and New York. Your right index finger is not, hopefully, at the same position of your his left eye. In fact, based on what we saw before, any object of finite mass is a distribution over a region, as discontinuous distributions are ruled out. An electron is an object of finite mass. Why should it be any different? In fact, shouldn't the burden of proof be on the claim that the electron is point-like, unlike any other finite mass object?

Part of the confusion seem to be the idea is that the electron is an elementary particle, and therefore should have no spatial extent. But that does not follow. Lack of spatial extend and lack of substructure are independent concepts. In fact: we can construct an infinitesimal object that is not elementary. Suppose you have dots constrained on the surface of a sphere. All their dynamics can be expressed by their polar and azimuthal angles. If we scale the sphere, the dynamics stays the same. We can shrink the whole sphere to a point: it is now a point-like object that is not elementary, as there is motion on its surface.

Not only point-like object don't have to be elementary, but it's not clear how they can be elementary. How can the whole universe be reduced to elementary elements that have no idea how many spatial dimensions are there? Mathematically: the set of points by itself does not define the topology. How is it supposed to work?

@@

The first payoff of accepting that an electron is a finite amount of material distributed in space is that this is more consistent with classical mechanics as we have defined it\footnote{It is also more consistent with general relativity.}. The second payoff is that some features of classical mechanics already explain part of the behavior that we thought was inherently quantum.

@@

Suppose we have a finite amount of classical material distributed over a single d.o.f. $a_m = \int \rho_\mathcal{c} dq^n \wedge dp_n$. The Shannon entropy associated with the distribution, which can be though as the amount of information required to identify a particle of the distribution, is finite as well. $I(\rho_\mathcal{c}) = \int \rho_\mathcal{c} \log (\rho_\mathcal{c})  dq^n \wedge dp_n = I_0$. TODO: prove in appendix.

As the state evolves into its final state, the amount of material remains the same: $a_m = \int \rho_{\hat{\mathcal{c}}} dq^n \wedge dp_n$. The final entropy will also remain the same since the entropy, like the distribution itself, is just another integrable function: $I(\rho_{\hat{\mathcal{c}}}) = I_0$.

@@

We now ask the following question: once we fix the entropy, what is the relationship between the spread of the marginal distributions along $q$ and $p$? Using Lagrange multipliers, we fix the total amount of material $a_m$ and the total entropy $I_0$ and find the distribution that minimizes the the product of the variance $\sigma_q^2 \sigma_p^2 = \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - \rho_{tot})\\
&+ \lambda_2(- \int \rho \ln \rho \, dqdp - H_0)\\
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\ &+ \lambda_1 + \lambda_2 \ln \rho + \lambda_2 ] dqdp = 0 \\
\rho = &\frac{\rho_{tot}}{ \sqrt{4 \pi ^2\sigma_q^2\sigma_p^2}} e^{-\frac{(q-\mu_q)^2(p-\mu_p)^2}{4\sigma_q^2\sigma_p^2}} \\
= &\frac{\rho_{tot}}{e^{H_0 - 1}} e^{-\frac{\pi ^2(q-\mu_q)^2(p-\mu_p)^2}{\exp (2(H_0 - 1))}}
\end{align*}
We find that the distribution that minimizes the spread is the product of two independent Gaussians. In quantum mechanics, the Gaussian wave packet is the state that minimizes uncertainty: this is the classical analogue.

As the distribution evolves, the entropy is conserved and therefore the product $\sigma_q^2 \sigma_p^2$ can never be less the one given by the Gaussian distribution of the same entropy. We have:
\begin{align*}
\sigma_q\sigma_p \geq \exp (H_0 - 1) / 2 \pi 
\end{align*}
This constitute the classical equivalent of the Heisenberg uncertainty principle.

The generalization to multiple d.o.f. is outside of the scope of this work. Note that this result is similar in spirit to the Gromov non-squeezing theorem and the symplectic camel, which are already generalized to multiple d.o.f., but the link here with information theory gives us further insight. In fact, it allows to understand the following picture.

Suppose the overall motion is deterministc and reversible but that each particle is undergoing non deterministic motion. Suppose $\rho$ is a distribution at equilibrium of such motion. While each particle moves randomly, the overall distribution $\rho$ stays the same. The non deterministic motion is effectively just reshuffling the particles within the same distribution.

Suppose now $\rho$ evolves adiabatically. At each instant we can assume it to be at equilibrium. At each instant the particles are reshuffled but the amount of entropy of the distribution remains the same. The overall evolution is still Hamiltonian, but the particle trajectories are now averages. The value of the entropy of the distribution is set by the non deterministic motion of the particles.

If we assumed that all materials were interacting with the same background that caused all particles to exhibit a random walk of the same magnitude, as defined by the information entropy, then the classical uncertainty principle would apply to all materials with the same constant. It would also be in line to address some of the problems mentioned in \ref{infinitesimal_reducibility}, as determinism and reversibility at all scales conceptually does not work.

It should now be clear that in classical mechanics we can already find some of the ingredients that are typically assumed to be characteristics of just quantum mechanics. Starting from well defined physical concepts is what allowed us to make such connection in simple and precise terms. Classical particles are not point-like objects, they are the limit of infinitesimal reductionism, and any finite amount of material is a distribution over phase space. Quantum particles are made of a finite amount of material, therefore will inherit some of their behavior from classical distributions. We are now ready to see where they get the remaining behavior.

@@

>>>>

As we saw in the previous sections, classical particles should not be understood as point-like objects of finite mass as this would open a series of problems and inconsistencies. The issues created by such misidentification are only amplified within the context of quantum mechanics. In fact, many issues preventing a satisfying intuitive picture stem from this confusion.

The problem goes along these line. One assumes, implicitly or explicitly, that quantum particles are point-like. But they don't always behave like one, they also behave like a wave, a distribution. During measurements, this distribution coincides experimentally with a probability distribution. So the state of a quantum particle is a probability distribution. Except that it self-interacts. So its a special quantum probability distribution as the quantum particle behaves like its in more than one place at once. We learn this picture and, sometimes begrudgingly, we get used to it.

Here we contend that this picture is not only confusing and unnecessary, it also prevents to understand fully the deep links between classical and quantum mechanics, in the same way that thinking of classical particles as point-like objects prevented to see a more holistic picture in classical mechanics.

The first thing to note is that all physical objects behave like they are in more than one place at once. The earth is both under Milano and New York. Your right index finger is not, hopefully, at the same position of your left eye. In fact, based on what we saw before, any object of finite mass is a distribution over a region, as discontinuous distributions are ruled out. An electron is an object of finite mass. Why should it be any different? In fact, if one claimed an electron to be point-like, he would be claiming an electron behaves in a different way than all other physical objects. Shouldn't \emph{he} have the burden of proof?

Part of the confusion seem to be the idea that, since the electron is an elementary particle, it should have no spatial extent. But that does not follow. Lack of spatial extend and lack of substructure are independent concepts. In fact: we can construct an infinitesimal object that is not elementary. Suppose we have a sphere with dots moving on its surface. All their dynamics can be expressed by their polar and azimuthal angles. If we scale the sphere, the dynamics stays the same. We can shrink the whole sphere to a point: it is now a point-like object that is not elementary, as there is motion on its surface.

Not only point-like object don't have to be elementary, but it's not clear how they can be elementary. How can the whole universe be reduced to  elements that have no idea how many spatial dimensions are there? Mathematically a set of points by itself does not define the dimension of a space: $\mathbbm{R}^n$ is the same order of infinity for any $n$. One needs a topology, which is an additional structure. If everything is just points, where does this come from?

@@

The first payoff of accepting that a quantum particle is a finite amount of material distributed in space is that this is more consistent with classical mechanics as we have defined it\footnote{It is also more consistent with general relativity.}. The second payoff is that some features of classical mechanics already explain part of the behavior that is usually thought as inherent to quantum mechanics.

@@

Suppose we have a finite amount of classical material distributed over a single d.o.f. $a_m = \int \rho_\mathcal{c} dq \wedge dp$. The Shannon entropy associated with the distribution, which can be though as the amount of information required to identify a particle of the distribution, is finite as well. $I(\rho_\mathcal{c}) = \int \rho_\mathcal{c} \log (\rho_\mathcal{c})  dq \wedge dp = I_0$. TODO: prove in appendix.

As the state evolves into its final state, the amount of material remains the same: $a_m = \int \rho_{\hat{\mathcal{c}}} dq \wedge dp$. The final entropy will also remain the same since the entropy, like the distribution itself, is just another integrable function: $I(\rho_{\hat{\mathcal{c}}}) = I_0$.

@@

We now ask the following question: once we fix the entropy, what is the relationship between the spread of the marginal distributions along $q$ and $p$? Using Lagrange multipliers, we fix the total amount of material $a_m$ and the total entropy $I_0$ and find the distribution that minimizes the the product of the variance $\sigma_q^2 \sigma_p^2 = \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - \rho_{tot})\\
&+ \lambda_2(- \int \rho \ln \rho \, dqdp - H_0)\\
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\ &+ \lambda_1 + \lambda_2 \ln \rho + \lambda_2 ] dqdp = 0 \\
\rho = &\frac{\rho_{tot}}{ \sqrt{4 \pi ^2\sigma_q^2\sigma_p^2}} e^{-\frac{(q-\mu_q)^2(p-\mu_p)^2}{4\sigma_q^2\sigma_p^2}} \\
= &\frac{\rho_{tot}}{e^{H_0 - 1}} e^{-\frac{\pi ^2(q-\mu_q)^2(p-\mu_p)^2}{\exp (2(I_0 - 1))}}
\end{align*}
We find that the distribution that minimizes the spread is the product of two independent Gaussians. Recall that in quantum mechanics, the Gaussian wave packet is the state that minimizes uncertainty: this is the classical analogue.

As the distribution evolves, the entropy is conserved and therefore the product $\sigma_q^2 \sigma_p^2$ can never be less than the one given by the Gaussian distribution of the same entropy. We have:
\begin{align*}
\sigma_q\sigma_p \geq \exp (I_0 - 1) / 2 \pi 
\end{align*}
This is strikingly similar to the Heisenberg uncertainty principle, except that the value depends on the initial state. This can be addressed with some extra assumptions.

Suppose the evolution of the distribution is deterministc and reversible but that each particle is undergoing non deterministic motion. Suppose $\rho$ is a distribution at equilibrium of such motion. While each particle moves randomly, the overall distribution $\rho$ stays the same: the non deterministic motion is effectively just reshuffling the particles within the same distribution. The value of the entropy of the distribution is set by the non deterministic motion of the particles: the stronger it is, the greater the entropy.

Suppose now $\rho$ evolves adiabatically. At each instant we can assume $\rho$ to be at equilibrium. At each instant the particles are reshuffled but the amount of entropy of the distribution remains the same. The overall evolution is still Hamiltonian, but the trajectories predicted by the framework are now the average motion of the particles.

If we assumed that all materials were interacting with the same background that caused all particles to exhibit a random walk of the same magnitude, as defined by the information entropy, then the spread of all distributions for all material will satisfy the same inequality. This is gives us the classical uncertainty principle: $\sigma_q\sigma_p \geq \exp (I_\varnothing - 1) / 2 \pi $ where $I_\varnothing$ is the entropy associated with the background.

Assuming such a background is not far fetched. In fact, it serves as a remainder of the issues mentioned in assumption \ref{infinitesimal_reducibility}: as determinism and reversibility at all scales conceptually does not work, there exists a minimum constant interaction to all particles: the system is not perfectly isolated. If the entropy of our distribution is comparable to the entropy introduced by the background, we are likely in the range where classical mechanics is not valid as the random motion of each particle is not negligible.

The generalization to multiple d.o.f. is outside of the scope of this work. Note, though, that this result is similar in spirit to the Gromov non-squeezing theorem and the symplectic camel, which are already generalized to multiple d.o.f.. It is the link with information theory that gives us further physical insight.

It should now be clear how in classical mechanics we can already find some of the ingredients that are typically assumed to be characteristics of quantum mechanics. It has its own flavor of anti-particles, CPT theorem and uncertainty principle. These are general features of an evolution that is deterministic, reversible and adiabatic (i.e. Hamiltonian dynamics). Quantum mechanics, as we'll see, just inherits these traits.

@@

We now focus on a different type of homogeneous material: one that is irreducible. That is, the state and the evolution of the whole does not tell us anything about the state and the evolution of the parts. As the internal dynamics is undetermined, we'll find that each irreducible element is suitably described by the standard framework of quantum mechanics.
